A Fully Abstract Game Semantics for Countable Nondeterminism

### 2018-04-07 cones/paper/lipics-v2018-authors/cones.tex
Made changes suggested by the seconnd reviewer.

We shall explore some of the problems relating to unbounded nondeterminism, and how they may be solved, using game semantics to give a fully abstract model of a simple stateful language -- Idealized Algol -- enhanced with a countable nondeterminism primitive.  
We begin with a pair of examples that will illustrate the lack of continuity, from a syntactic point of view.  
Let $\nat$ be our natural number type and consider a sequence of functions ${<}n\from\nat\to\nat$, where ${<}m\;k$ evaluates to $0$ if $k<n$ and diverges otherwise.
In that case, the least upper bound of the ${<}n$ is the function that combines all their convergent behaviours; i.e., the function $\lambda k.k;\0$ that evaluates its input and then returns $0$.
If $\wn\from\nat$ is an unbounded nondeterminism machine, then function application to $\wn$ is not continuous; indeed, ${<}m\;\wn$ always has the possibility of divergence -- since $\wn$ may evaluate to $m+1$, say.  
But $(\lambda k.k;\0)\;\wn$ always converges to $0$.

@@

We describe a simple type theory and operational semantics for Idealized Algol with countable nondeterminism.
This is similar to the approach adopted in \cite{mcCHFiniteND}, which extends Idealized Algol with \emph{finite} nondeterminism.  
The types of our language are defined inductively as follows:
\[
  T \Coloneqq \nat \mid \com \mid \Var \mid T \to T\,.
  \]
Meanwhile, the terms are those given in \cite{SamsonGuyIAPassive}, together with the nondeterministic choice:
\begin{IEEEeqnarray*}{RCL}
  M & \Coloneqq & x \mid \lambda x . M \mid M\;M \mid \Y_T \mid \\
  && \n \mid \skipp \mid \suc \mid \pred \mid \\
  && \IfO \mid \blank;\blank \mid \blank \coloneqq \blank \mid \\
  && \deref \mid \neww_T \mid \mkvar \mid \wn\,.
\end{IEEEeqnarray*}

@@

\begin{figure*}
  \begin{mathpar}
    \inferrule*{ }{\ia s {(\lambda x.M)\;N} \opto \ia s {M[N/x]}}
    \and
    \inferrule*{ }{\ia s {\Y_T M}\opto \ia s {M (\Y_T M)}}
    \and
    \inferrule*{ }{\ia s {\suc \n}\opto\ia s {\n + \mathsf{1}}}
    \and
    \inferrule*{ }{\ia s {\pred \n}\opto\ia s {\0 \sqcup (\n - \mathsf{1})}}
    \and
    \inferrule*{ }{\ia s {\IfO \0 M N} \opto \ia s M}
    \and
    \inferrule*{ }{\ia s {\IfO (\n + \mathsf{1}) M N} \opto \ia s N}
    \and
    \inferrule*{ }{\ia s {\deref (\mkvar M N)} \opto \ia s M}
    \and
    \inferrule*{ }{\ia s {(\mkvar M N) \coloneqq L} \opto \ia s {N\;L}}
    \and
    \inferrule*{ }{\ia s {v \coloneqq n} \opto \ia {\stup s v n} \skipp}
    \and
    \inferrule*[left={$s(v)=n$}]{ }{\ia s {\deref v} \opto \ia s n}
    \and
    \inferrule*{ }{\ia s {\skipp;M} \opto \ia s M}
    \and
    \inferrule*{ }{\ia s {\new_T \lambda v.M}\opto \ia {\stup s v 0} M}
    \and
    \inferrule*{\ia s M \opto \ia s {M'}}{\ia s {E[M]} \opto \ia s {E[M']}}
    \and
    \inferrule*[right=$n\in\bN$]{ }{\ia s \wn \opto \ia s \n}
  \end{mathpar}
  \caption{Small-step operational semantics for Idealized Algol with countable nondeterminism}
  \label{fig:ia-os}
\end{figure*}/

@@

Let $A,B$ be games.  
Then we may define games $A\times B$, $A\tensor B$, $A\implies B$ and $\oc A$ as in \cite{SamsonGuyIAPassive}.  
As an example, we give the definition of $A\implies B$:
\begin{IEEEeqnarray*}{RCL}
  M_{A\implies B} & \quad=\quad & M_A + M_B\,. \\
  \lambda_{A\implies B} & = & [\neg\circ\lambda_A,\lambda_B]\,.\\
  *\ts_{A\implies B} n & \Leftrightarrow & * \ts_B m\,.\\[1.0ex]
  m \ts_{A\implies B} n & \Leftrightarrow & \mbox{\pbox\textwidth{$m \ts_A n$ or $m\ts_B n$ \\ or (for $m\neq *$) $ * \ts_B m$ and $* \ts_A n$\,.}} \\[1.0ex]
  P_{A\implies B} & = & \{s\in L_{A\implies B}\suchthat s\vert_A\in P_A\text{ and }s\vert_B\in P_B\}\,.
\end{IEEEeqnarray*}

>>>>

We shall explore some of the problems relating to unbounded nondeterminism, and how they may be solved, using game semantics to give a fully abstract model of a simple stateful language -- Idealized Algol -- enhanced with a countable nondeterminism primitive.  
We begin with a pair of examples that will illustrate the lack of continuity, from a syntactic point of view.  
Let $\nat$ be our natural number type and consider a sequence of functions ${<}n\from\nat\to\nat$, where ${<}n\;k$ evaluates to $0$ if $k<n$ and diverges otherwise.
In that case, the least upper bound of the ${<}n$ is the function that combines all their convergent behaviours; i.e., the function $\lambda k.k;\0$ that evaluates its input and then returns $0$.
If $\wn\from\nat$ is an unbounded nondeterminism machine, then function application to $\wn$ is not continuous; indeed, ${<}m\;\wn$ always has the possibility of divergence -- since $\wn$ may evaluate to $m+1$, say.  
But $(\lambda k.k;\0)\;\wn$ always converges to $0$.

@@

We describe a simple type theory and operational semantics for Idealized Algol with countable nondeterminism.
This is similar to the approach adopted in \cite{mcCHFiniteND}, which extends Idealized Algol with \emph{finite} nondeterminism.  
The types of our language are defined inductively as follows:
\[
  T \Coloneqq \nat \mid \com \mid \Var \mid T \to T\,.
  \]
Meanwhile, the terms are those given in \cite{SamsonGuyIAPassive}, together with the nondeterministic choice:
\begin{IEEEeqnarray*}{RCL}
  M & \Coloneqq & x \mid \lambda x . M \mid M\;M \mid \Y_T \mid \\
  && \n \mid \skipp \mid \suc \mid \pred \mid \\
  && \IfO \mid \blank;\blank \mid \blank \coloneqq \blank \mid \\
  && \deref\footnote{That is, variable \deref{}ccess.} \mid \neww_T \mid \mkvar \mid \wn\,.
\end{IEEEeqnarray*}

@@

\begin{figure*}
  \begin{mathpar}
    \inferrule*{ }{\ia s {(\lambda x.M)\;N} \opto \ia s {M[N/x]}}
    \and
    \inferrule*{ }{\ia s {\Y_T M}\opto \ia s {M (\Y_T M)}}
    \and
    \inferrule*{ }{\ia s {\suc \n}\opto\ia s {\n + \mathsf{1}}}
    \and
    \inferrule*{ }{\ia s {\pred \n}\opto\ia s {\0 \sqcup (\n - \mathsf{1})}}
    \and
    \inferrule*{ }{\ia s {\IfO \0 M N} \opto \ia s M}
    \and
    \inferrule*{ }{\ia s {\IfO (\n + \mathsf{1}) M N} \opto \ia s N}
    \and
    \inferrule*{ }{\ia s {\deref (\mkvar M N)} \opto \ia s M}
    \and
    \inferrule*{ }{\ia s {(\mkvar M N) \coloneqq L} \opto \ia s {N\;L}}
    \and
    \inferrule*{ }{\ia s {v \coloneqq n} \opto \ia {\stup s v n} \skipp}
    \and
    \inferrule*[left={$s(v)=n$}]{ }{\ia s {\deref v} \opto \ia s n}
    \and
    \inferrule*{ }{\ia s {\skipp;M} \opto \ia s M}
    \and
    \inferrule*{ }{\ia s {\new_T \lambda v.M}\opto \ia {\stup s v 0} M}
    \and
    \inferrule*{\ia s M \opto \ia s {M'}}{\ia s {E[M]} \opto \ia s {E[M']}}
    \and
    \inferrule*[right=$n\in\bN$]{ }{\ia s \wn \opto \ia s \n}
  \end{mathpar}
  \caption{Small-step operational semantics for Idealized Algol with countable nondeterminism}
  \label{fig:ia-os}
\end{figure*}

@@

Let $A,B$ be games.  
Then we may define games $A\times B$, $A\tensor B$, $A\implies B$ and $\oc A$ as in \cite{SamsonGuyIAPassive}.  
As an example, we give the definition of $A\implies B$:
\begin{IEEEeqnarray*}{RCL}
  M_{A\implies B} & \quad=\quad & M_A + M_B\,. \\
  \lambda_{A\implies B} & = & [\neg\circ\lambda_A,\lambda_B]\,.\\
  *\ts_{A\implies B} n & \Leftrightarrow & * \ts_B n\,.\\[1.0ex]
  m \ts_{A\implies B} n & \Leftrightarrow & \mbox{\pbox\textwidth{$m \ts_A n$ or $m\ts_B n$ \\ or (for $m\neq *$) $ * \ts_B m$ and $* \ts_A n$\,.}} \\[1.0ex]
  P_{A\implies B} & = & \{s\in L_{A\implies B}\suchthat s\vert_A\in P_A\text{ and }s\vert_B\in P_B\}\,.
\end{IEEEeqnarray*}
